---
title: "22q_longitudinal_gamm"
author: "charlie schleifer"
date: "5/10/2022"
output: html_document
---
## Overview
This script uses Generalized Additive Mixed Models (GAMMs) to model non-parametric age trajectories for thalamocortical functional network connectivity in 22qDel patients and typical controls. A random effects implementation of the ComBat algorithm (longCombat) is used to harmonize data from multiple sites.  

## Set up workspace
ciftiTools is an R package for analysis and plotting of CIFTI dscalar, dlabel, and dtseries files:
https://htmlpreview.github.io/?https://github.com/mandymejia/ciftiTools/blob/master/vignettes/ciftiTools_vignette.html
many ciftiTools functions require connectome workbench to be downloaded and installed locally:
https://www.humanconnectome.org/software/get-connectome-workbench
```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
# clear workspace
rm(list = ls(all.names = TRUE))

# use SSHFS to mount hoffman2 server (download SSHFS for mac: https://osxfuse.github.io/)
# TODO: set hoffman2 username
uname <- "schleife"
# set local path to mount server
hoffman <- "~/Desktop/hoffman_mount"
# create directory if needed 
if(!file.exists(hoffman)){dir.create(hoffman)}
# make string to run as system command
mntcommand <- paste0("umount -f ", hoffman,"; sshfs ",uname,"@hoffman2.idre.ucla.edu:/u/project/cbearden/data ",hoffman)
# if hoffman directory is empty, use system command and sshfs to mount server, if not empty assume already mounted and skip
if(length(list.files(hoffman)) == 0){system(mntcommand)}else{print(paste(hoffman,"is not empty...skipping SSHFS step"))}

# install longCombat for first time
#install.packages("devtools")
#devtools::install_github("jcbeer/longCombat")

# list packages to load
# ciftiTools dependency rgl may need XQuartz installed in order to visualize surfaces
#packages <- c("conflicted", "here", "magrittr", "mgcv", "gratia", "lme4", "lmerTest", "invgamma", "longCombat", "ciftiTools", "readxl", "dplyr", "data.table", "DescTools","tableone", "tibble", "reshape2", "viridis", "scico", "ggplot2", "gridExtra", "ggpubr","stringr")
packages <- c("conflicted", "here", "magrittr", "readxl", "dplyr", "data.table", "DescTools","tableone", "tibble", "reshape2", "viridis", "scico", "ggplot2", "gridExtra", "ggpubr","stringr", "parallel")

# install packages if not yet installed
all_packages <- rownames(installed.packages())
installed_packages <- packages %in% all_packages
if (any(installed_packages == FALSE)){install.packages(packages[!installed_packages])}

# load packages
invisible(lapply(packages, library, character.only = TRUE))

# use the filter function from dplyr, not stats
conflict_prefer("filter", "dplyr")

# get path to project repo directory
project <- here()
# for supplemental scripts, keep project dir as main repo
project <- dirname(project)
print(paste("Project directory:", project))

# set up connectome workbench path for ciftiTools
# https://www.humanconnectome.org/software/get-connectome-workbench
# local wbpath (edit this path if workbench is installed in another location, e.g. on hoffman: /u/project/CCN/apps/hcp/current/workbench/bin_rh_linux64/)
# TODO: edit if necessary
#wbpath <- "/Applications/workbench/bin_macosx64/"
#ciftiTools.setOption("wb_path", wbpath)

# load rgl for ciftiTools visualization
# may require XQartz v2.8.1 to be installed locally
#if(!require('rgl', quietly=TRUE)){install.packages('rgl')}
#rgl::setupKnitr()
#rgl::rgl.open(); rgl::rgl.close()
```


## Load individual TC connectivity CSVs
computed by 22q_multisite_networkTC_save_individual.R and saved as a CSV with one value per network representing the z-transformed pearson correlation between signals in the thalamic and cortical subsets of that network
```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
# paths to sessions directories
#trio_dir <- file.path(hoffman,"22q/qunex_studyfolder/sessions")
#prisma_dir <- file.path(hoffman,"22qPrisma/qunex_studyfolder/sessions")
file_dir <- "~/Dropbox/PhD/bearden_lab/22q/analyses/files_from_hoffman/sessions/"

# get list of sessions
all_sessions <- list.files(file_dir)

#trio_sessions <- list.files(trio_dir,pattern="Q_[0-9]")
#prisma_sessions <- list.files(prisma_dir,pattern="Q_[0-9]")
# exclude Q_0390_09302019 for now due to no AP BOLD; test excluding "Q_0321_03272017","Q_0334_12012016" to ensure restingAP* vs resting* prisma are the same
#exclude_sessions <- c("Q_0390_09302019","Q_0321_03272017","Q_0334_12012016")
#exclude_sessions <- c("Q_0390_09302019","Q_0477_01052022","Q_0484_01042022","Q_0508_06232022","Q_0519_05312022","Q_0520_06012022","Q_0521_05202022","Q_0525_06072022","Q_0526_06242022","Q_0527_07112022","Q_0528_07202022","Q_0529_07202022","Q_0541_07182022","Q_0549_10182022","Q_0561_11032022","Q_0568_10252022")

# exclude only Q_0390_09302019 (missing AP BOLD)
exclude_sessions <- "Q_0390_09302019"
#prisma_sessions <- prisma_sessions[! prisma_sessions %in% exclude_sessions]
#all_sessions <- c(trio_sessions,prisma_sessions)
all_sessions <- all_sessions[! all_sessions %in% exclude_sessions]

# function to read parcellated connectivity results and add columns for roi pair name, site, and ID 
read_bparc_results <- function(sdir, fname, sesh){
  print(sesh)
  output <- as.data.frame(read.csv(file.path(sdir,sesh,"images/functional",fname)))
  output$MRI_S_ID <- sesh
  return(output)
}

# file name to look for
bparc_name_trio_gsr <- "resting_fc_matrix_Atlas_s_hpss_res-mVWMWB1d_lpss_CABNP_between_parcel.csv"
bparc_name_trio_nogsr <- "resting_fc_matrix_Atlas_s_hpss_res-mVWM1d_lpss_CABNP_between_parcel.csv"
bparc_name_prisma_gsr <- "restingAP_fc_matrix_Atlas_s_hpss_res-mVWMWB1d_lpss_CABNP_between_parcel.csv"
bparc_name_prisma_nogsr <- "restingAP_fc_matrix_Atlas_s_hpss_res-mVWM1d_lpss_CABNP_between_parcel.csv"

# read for trio and prisma then combine
all_bparc_gsr <- lapply(all_sessions, function(s) read_bparc_results(sesh=s,sdir=file_dir,fname=dir(file.path(file_dir,s,"images/functional/") ,pattern="resting.*_fc_matrix_Atlas_s_hpss_res-mVWMWB1d_lpss_CABNP_between_parcel.csv"))) %>% do.call(rbind,.) %>% as.data.frame

all_bparc_nogsr <- lapply(all_sessions, function(s) read_bparc_results(sesh=s,sdir=file_dir,fname=dir(file.path(file_dir,s,"images/functional/") ,pattern="resting.*_fc_matrix_Atlas_s_hpss_res-mVWM1d_lpss_CABNP_between_parcel.csv"))) %>% do.call(rbind,.) %>% as.data.frame

# add column with roi1 and roi2 numbers
all_bparc_gsr$rois <- paste0(all_bparc_gsr$roi_1,"_",all_bparc_gsr$roi_2)
all_bparc_nogsr$rois <- paste0(all_bparc_nogsr$roi_1,"_",all_bparc_nogsr$roi_2)

# read motion data
movement <- read.csv("~/Dropbox/PhD/bearden_lab/22q/analyses/files_from_hoffman/22q_multisite_fd_mean.csv", header=TRUE)

# cast to wide for combat
#setDT(all_tc)
#all_tc_wide <- reshape2::dcast(all_tc, MRI_S_ID + site ~ NETWORK, value.var="TC_Fz") 
```


## load sistat data and get lists of scans to use
all sistat tables should be exported as CSVs into a single directory
the next several chunks deal with reading, cleaning and annotating the data exported from sistat, and then age matching
the hcs sample is younger than del due to a large amount of very young hcs subjects. plan is to match samples by using followup timepoints rather than baseline for some younger participants, and dropping several older del subjects, and younger hcs subjects (prioritizing dropping subjects with worse motion stats when possible)
```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
# set location of directory with ucla sistat CSVs
csvdir_ucla <- file.path(project,"demographics/ucla_sistat")

# get list of files_ucla in directory
files_ucla <- list.files(csvdir_ucla)
fpaths <- lapply(files_ucla, function(file) paste(csvdir_ucla,file,sep="/"))

# clean names
fnames <- gsub(".csv","",files_ucla)
fnames <- gsub("Re22Q_","",fnames)
fnames <- gsub("Form_","",fnames)
fnames <- gsub("Qry_","",fnames)

# read all, set to na: "-9999", "-9998","." 
input_all_ucla <- lapply(fpaths, read.csv, header=T, na.strings=c(".","-9999","-9998"), strip.white=T, sep=",")
names(input_all_ucla) <- fnames
df_all_ucla <- lapply(input_all_ucla, function(x) data.frame(x))

# subset demo_mri for used scans
ucla_demo <- df_all_ucla$demo_mri %>% filter(MRI_S_ID %in% all_sessions)

# remove "FAMILY MEMBER" designation from subject identity
ucla_demo$SUBJECT_IDENTITY <- ucla_demo$SUBJECT_IDENTITY %>% sub("FAMILY MEMBER","",.) %>% sub(",","",.) %>% trimws(which="both") %>% as.factor
# change sex coding from 0/1 to F/M and set to factor
ucla_demo$SEX <- factor(ucla_demo$SEX,levels=c(0,1),labels=c("F","M"))
```

# TEMPORARY
read temporary csv with several subjects not yet in sistat
```{r}
# TODO: this chunk is temporary until sistat is updated 
# TODO: note: q_0526 sex was "na" in original sheet, changed to M because combat can't have NAs
# read new data
temp_demo <- read_xlsx(file.path(project,"demographics/temporary/sMRI_demo_info_forCharlie.xlsx"), col_names=TRUE,na="",trim_ws = TRUE)

# make empty demographics data frame to add new data to
demo_add <- ucla_demo[1:nrow(temp_demo),]
demo_add[,] <- NA
demo_add$SUBJECTID <- temp_demo$`Subject ID`
demo_add$SUBJECT_IDENTITY <- temp_demo$Diagnosis
demo_add$MRI_S_ID <- temp_demo$`MRI ID`
demo_add$SEX <- as.factor(temp_demo$Sex)
demo_add$AGE <- temp_demo$Age
demo_add$AGEMONTH <- temp_demo$Age*12
demo_add$CONVERTEDVISITNUM <- 2

# append to ucla demo
ucla_demo <- rbind(ucla_demo,demo_add)
```
 
continue regular steps
```{r}
# manually fix missing sex for Q_0381_09102019
# TODO: fix in sistat and re-export
ucla_demo[which(ucla_demo$MRI_S_ID == "Q_0381_09102019"),"SEX"] <- "F"

# set race=NA to 7 (unknown)
ucla_demo$RACE[is.na(ucla_demo$RACE)] <- 7
# set race as factor 1=American Indian/Alaska Native; 2=Asian; 3=Native Hawaiian/Pacific Islander; 4=Black or African American; 5=White; 6=Multiple; 7=Unknown
ucla_demo$RACE <- factor(ucla_demo$RACE,levels=c(1:7),labels=c("1_Native_American","2_Asian","3_Pacific_Island","4_Black","5_White","6_Multiple","7_Unknown"))
# ethnicity as factor with 0=N 1=Y
ucla_demo$HISPANIC[is.na(ucla_demo$HISPANIC)] <- "Unknown"
ucla_demo$HISPANIC <- factor(ucla_demo$HISPANIC,levels=c(0,1,"Unknown"),labels=c("N","Y","Unknown"))
# get more accurate age with AGEMONTH/12
ucla_demo$AGE <- as.numeric(ucla_demo$AGEMONTH)/12 

# function to add column to code timepoints relative to sample used (i.e. if visit 1 and 1.12 missing, then 1.24 is baseline)
# trio/prisma coded as T/P-visit_n where T-1 would be the subject's first trio scan and P-1 the first prisma, P-2 the second...
# function should be applied to the indicies of rows (r) in a subset of demo_mri
gettp <- function(r, df){
  sub <- df$SUBJECTID[[r]]
  visit <- df$CONVERTEDVISITNUM[[r]]
  all_visits <- df$CONVERTEDVISITNUM[which(df$SUBJECTID == sub)] %>% sort
  n_visits <- length(all_visits)
  nt_visits <-length(which(all_visits < 2))
  np_visits <- length(which(all_visits >= 2))
  visit_index <- which(all_visits == visit)
  if (visit < 2){
    label=paste("T-",visit_index,sep="")
  }else if (visit >= 2){
    p_visits <- all_visits[which(all_visits >= 2)] %>% sort
    p_visit_index <- which(p_visits == visit)
    label=paste("P-",p_visit_index,sep="")
  }
  return(c(sub,visit,label,n_visits,nt_visits,np_visits,visit_index))
}

# get timepoints
timepoints <- sapply(1:nrow(ucla_demo),function(r) gettp(r,ucla_demo)) %>% t %>% as.data.frame
colnames(timepoints) <- c("SUBJECTID","CONVERTEDVISITNUM","converted_timepoint","n_timepoints","n_trio","n_prisma","visit_index")
ucla_demo_tp <- cbind(ucla_demo,timepoints[,3:7])
ucla_demo_tp$visit_index %<>% as.factor

# subset to under max age limit (22 years old)
ucla_demo_tp_agelim <- filter(ucla_demo_tp, ucla_demo_tp$AGE < 22)

# subset to hcs del
ucla_demo_hcs_del <- ucla_demo_tp_agelim %>% filter(SUBJECT_IDENTITY=="CONTROL" | SUBJECT_IDENTITY =="PATIENT-DEL")

# remove unused factor levels
ucla_demo_hcs_del %<>% droplevels
```

All timepoints, pre-matching demographics summary
```{r}
demo_summary <- CreateTableOne(data=ucla_demo_hcs_del,vars=c("AGE","SEX"),strata="SUBJECT_IDENTITY",addOverall=F)
print(demo_summary, showAllLevels=T)
```

Baseline pre-matching summary
```{r}
demo_summary_bl <- CreateTableOne(data=filter(ucla_demo_hcs_del, ucla_demo_hcs_del$visit_index == 1),vars=c("AGE","SEX"),strata="SUBJECT_IDENTITY",addOverall=F)
print(demo_summary_bl)
```

```{r}
# get baseline sample
ucla_demo_hcs_del_bl <- filter(ucla_demo_hcs_del, ucla_demo_hcs_del$visit_index == 1)

# filter connectivity data for baseline sample
all_bparc_gsr_bl <- filter(all_bparc_gsr, MRI_S_ID %in% ucla_demo_hcs_del_bl$MRI_S_ID)
all_bparc_nogsr_bl <- filter(all_bparc_nogsr, MRI_S_ID %in% ucla_demo_hcs_del_bl$MRI_S_ID)

# unique roi pairs
roipairs <- as.vector(unique(all_bparc_gsr$rois))

# set stuff we don't need to null
all_bparc_gsr <- NULL
all_bparc_nogsr <- NULL

# for each roi pair, compute correlation with fd_mean across subjects
qcfc_roi <- function(df, roi, movement){
  print(roi)
  merged <- merge(x=filter(df, rois==roi), y=movement, by="MRI_S_ID")
  out <- cor(x=merged$pearson_r_Fz, y=merged$fd_mean)
  return(out)
}
qcfc_gsr_bl <- lapply(roipairs, function(r)qcfc_roi(df=all_bparc_gsr_bl, roi=r, movement=movement)) %>% do.call(rbind,.) %>% as.data.frame
colnames(qcfc_gsr_bl) <- "qcfc_gsr"

qcfc_nogsr_bl <- lapply(roipairs, function(r)qcfc_roi(df=all_bparc_nogsr_bl, roi=r, movement=movement)) %>% do.call(rbind,.) %>% as.data.frame
colnames(qcfc_nogsr_bl) <- "qcfc_nogsr"

qcfc <- cbind(qcfc_gsr_bl, qcfc_nogsr_bl)
qcfc$roi_pair <- roipairs


```


read qcfc computed on hoffman
```{r}
qcfc_all <- read.csv(file.path(hoffman,"22q/qunex_studyfolder/analysis/fcMRI/qcfc_22q_baseline_all.csv"), header = TRUE)

t.test(x=qcfc_all$qcfc_gsr, y=qcfc_all$qcfc_nogsr, alternative = "two.sided", paired = TRUE)

mean(qcfc_all$qcfc_gsr)
mean(qcfc_all$qcfc_nogsr)

qcfc_long <- data.table::melt(qcfc_all, measure.vars=c("qcfc_gsr","qcfc_nogsr"), id.vars="roi_pair")
qcfc_long$variable <- gsub("qcfc_gsr","GSR",qcfc_long$variable)
qcfc_long$variable <- gsub("qcfc_nogsr","No GSR",qcfc_long$variable)

qcfc_plot <- ggplot(qcfc_long, aes(y=value,x=variable,fill=variable))+
  geom_boxplot(outlier.shape =21, outlier.fill="white", outlier.alpha = 0.7)+
  #scale_fill_manual(values=c("green","purple"))+
  scale_fill_viridis(discrete = TRUE)+
  stat_compare_means(method="wilcox.test", label.y=0.65)+
  #stat_compare_means(method="t.test", label="p.signif", label.y=0.75)+
  ggtitle("Mean Correlation between FC and Movement (QC-FC)")+
  xlab("Global Signal Regression status")+
  #scale_x_discrete(labels=c("GSR", "No GSR"))+
  ylab("Pearson's r")+
  theme_classic()

ggsave(plot=qcfc_plot, file=file.path(project,"figures/sensitivity_analyses/qcfc_gsr.svg"), device="svg")
ggsave(plot=qcfc_plot, file=file.path(project,"figures/sensitivity_analyses/qcfc_gsr.png"), device="png")
```






