---
title: "22qDel Longitudinal Thalamocortical GAMM"
author: "Charles Schleifer"
date: "6/23/2022"
output: github_document
---
### Overview
This script uses Generalized Additive Mixed Models (GAMMs) to model non-parametric age trajectories for thalamocortical functional network connectivity in 22qDel patients and typical controls. A random effects implementation of the ComBat algorithm (longCombat) is used to harmonize data from multiple sites.  

### Set up workspace
ciftiTools is an R package for analysis and plotting of CIFTI dscalar, dlabel, and dtseries files:
https://htmlpreview.github.io/?https://github.com/mandymejia/ciftiTools/blob/master/vignettes/ciftiTools_vignette.html
many ciftiTools functions require connectome workbench to be downloaded and installed locally:
https://www.humanconnectome.org/software/get-connectome-workbench
```{r message=FALSE, warning=FALSE, paged.print=FALSE, results='hide'}
# clear workspace
rm(list = ls(all.names = TRUE))

# use SSHFS to mount hoffman2 server (download SSHFS for mac: https://osxfuse.github.io/)
# TODO: set hoffman2 username
uname <- "schleife"
# set local path to mount server
hoffman <- "~/Desktop/hoffman_mount"
# create directory if needed 
if(!file.exists(hoffman)){dir.create(hoffman)}
# make string to run as system command
mntcommand <- paste0("umount -f ", hoffman,"; sshfs ",uname,"@hoffman2.idre.ucla.edu:/u/project/cbearden/data ",hoffman)
# if hoffman directory is empty, use system command and sshfs to mount server, if not empty assume already mounted and skip
if(length(list.files(hoffman)) == 0){system(mntcommand)}else{print(paste(hoffman,"is not empty...skipping SSHFS step"))}

# install longCombat for first time
#install.packages("devtools")
#devtools::install_github("jcbeer/longCombat")

# list packages to load
# ciftiTools dependency rgl may need XQuartz installed in order to visualize surfaces
packages <- c("conflicted", "here", "magrittr", "mgcv", "gratia", "lme4", "lmerTest", "invgamma", "longCombat", "ciftiTools", "readxl", "dplyr", "data.table", "DescTools","tableone", "tibble", "reshape2", "viridis", "scico", "ggplot2", "gridExtra", "ggpubr","stringr")

# install packages if not yet installed
all_packages <- rownames(installed.packages())
installed_packages <- packages %in% all_packages
if (any(installed_packages == FALSE)){install.packages(packages[!installed_packages])}

# load packages
invisible(lapply(packages, library, character.only = TRUE))

# use the filter function from dplyr, not stats
conflict_prefer("filter", "dplyr")

# get path to project repo directory
project <- here()
print(paste("Project directory:", project))

# set up connectome workbench path for ciftiTools
# https://www.humanconnectome.org/software/get-connectome-workbench
# local wbpath (edit this path if workbench is installed in another location, e.g. on hoffman: /u/project/CCN/apps/hcp/current/workbench/bin_rh_linux64/)
# TODO: edit if necessary
wbpath <- "/Applications/workbench/bin_macosx64/"
ciftiTools.setOption("wb_path", wbpath)

# load rgl for ciftiTools visualization
# may require XQartz v2.8.1 to be installed locally
if(!require('rgl', quietly=TRUE)){install.packages('rgl')}
rgl::setupKnitr()
rgl::rgl.open(); rgl::rgl.close()
```

### Load CAB-NP atlas
RSN atlas of whole cortex and subcortex to use for thalamus striatum connectivity analysis. Atlas can be downloaded here: https://github.com/ColeLab/ColeAnticevicNetPartition
load key for Ji parcels/networks
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
ji_key <- read.table(file.path(project,"CAB-NP/CortexSubcortex_ColeAnticevic_NetPartition_wSubcorGSR_parcels_LR_LabelKey.txt"),header=T)
ji_net_keys <- ji_key[,c("NETWORKKEY","NETWORK")] %>% distinct %>% arrange(NETWORKKEY)
print(ji_net_keys)

# read cifti with subcortical structures labeled 
xii_Ji_parcel <- read_cifti(file.path(project,"CAB-NP/CortexSubcortex_ColeAnticevic_NetPartition_wSubcorGSR_parcels_LR.dscalar.nii"), brainstructures = "all")
xii_Ji_network <- read_cifti(file.path(project,"CAB-NP/CortexSubcortex_ColeAnticevic_NetPartition_wSubcorGSR_netassignments_LR.dscalar.nii"), brainstructures = "all")

#view_xifti_volume(xii_Ji_parcel,colors="viridis",title="parcels",cex.title=1.3)
#view_xifti_volume(xii_Ji_network,colors="Paired",title="networks",cex.title=1.3)
```

### Load individual TC connectivity CSVs
Computed by 22q_multisite_networkTC_save_individual.R and saved as a CSV with one value per network representing the z-transformed pearson correlation between signals in the thalamic and cortical subsets of that network
```{r message=FALSE, warning=FALSE, paged.print=FALSE, results='hide'}
# paths to sessions directories
trio_dir <- file.path(hoffman,"22q/qunex_studyfolder/sessions")
prisma_dir <- file.path(hoffman,"22qPrisma/qunex_studyfolder/sessions")

# get list of sessions
trio_sessions <- list.files(trio_dir,pattern="Q_[0-9]")
prisma_sessions <- list.files(prisma_dir,pattern="Q_[0-9]")
# exclude Q_0390_09302019 for now due to no AP BOLD; test excluding "Q_0321_03272017","Q_0334_12012016" to ensure restingAP* vs resting* prisma are the same
#exclude_sessions <- c("Q_0390_09302019","Q_0321_03272017","Q_0334_12012016")
#exclude_sessions <- c("Q_0390_09302019","Q_0477_01052022","Q_0484_01042022","Q_0508_06232022","Q_0519_05312022","Q_0520_06012022","Q_0521_05202022","Q_0525_06072022","Q_0526_06242022","Q_0527_07112022","Q_0528_07202022","Q_0529_07202022","Q_0541_07182022","Q_0549_10182022","Q_0561_11032022","Q_0568_10252022")

# exclude only Q_0390_09302019 (missing AP BOLD)
exclude_sessions <- "Q_0390_09302019"
prisma_sessions <- prisma_sessions[! prisma_sessions %in% exclude_sessions]
all_sessions <- c(trio_sessions,prisma_sessions)

# function to read thalamocortical results and add columns for roi pair name, site, and ID 
read_tc_results <- function(sdir, fname, sesh, site){
  print(sesh)
  input <- read.csv(file.path(sdir,sesh,"images/functional",fname))
  session <- rep(sesh, times=nrow(input)) %>% as.data.frame
  site <- rep(site, times=nrow(input)) %>% as.data.frame
  new_cols <- cbind(session,site)
  colnames(new_cols) <- c("MRI_S_ID","site")
  output <- cbind(input,new_cols)
  return(output)
}

# file name to look for
tc_name_trio <- "resting_network_TCC_matrix_Atlas_s_hpss_res-mVWMWB1d_lpss_CABNP.csv"
tc_name_prisma <- "restingAP_network_TCC_matrix_Atlas_s_hpss_res-mVWMWB1d_lpss_CABNP.csv"
#tc_name_trio <- "resting_fc_network_Thal_Cortex_Atlas_s_hpss_res-mVWM1d_lpss_CABNP.csv"
#tc_name_prisma <- "restingAP_fc_network_Thal_Cortex_Atlas_s_hpss_res-mVWM1d_lpss_CABNP.csv"

# read for trio and prisma then combine
trio_tc <- lapply(trio_sessions, function(s) read_tc_results(sesh=s,site="trio",sdir=trio_dir,fname=tc_name_trio)) %>% do.call(rbind,.) %>% as.data.frame
prisma_tc <- lapply(prisma_sessions, function(s) read_tc_results(sesh=s,site="prisma",sdir=prisma_dir,fname=tc_name_prisma)) %>% do.call(rbind,.) %>% as.data.frame
#all_tc <- rbind(trio_tc,prisma_tc) %>% filter(!is.na(TC_Fz))
all_tc <- rbind(trio_tc,prisma_tc)
all_tc$pearson_r_Fz %<>% as.numeric

# replaces dashes with underscores for network names
#all_tc$NETWORK <- all_tc$NETWORK %>% gsub("-","_",.)

# cast to wide for combat
#setDT(all_tc)
#all_tc_wide <- reshape2::dcast(all_tc, MRI_S_ID + site ~ NETWORK, value.var="TC_Fz") 
```

for each subject, test if a given network in the thalamus is most highly correlated with that same cortical network
```{r message=FALSE, warning=FALSE, paged.print=FALSE, results='hide'}

# function to test a specific network and subject
net_rank <- function(all_data,  sesh, network){
  sub_data <- filter(all_data, MRI_S_ID==sesh)
  # get connectivity for thalamic and cortical network == chosen network
  chosen_net <- filter(sub_data, Thalamus==network & Cortex == network)$pearson_r_Fz 
  # get connectivity from that thalamic network to all other networks
  other_nets <- filter(sub_data, Thalamus==network & Cortex != network)$pearson_r_Fz %>% as.vector
  # test if chosen network is highest
  higher <- all(other_nets < chosen_net)
  out <- data.frame(MRI_S_ID=sesh, thal_net=network, highest=higher)
  return(out)
}

all_som_highest <- lapply(unique(all_tc$MRI_S_ID), function(x) net_rank(sesh=x, network="Somatomotor", all_data=all_tc)) %>% do.call(rbind,.) %>% as.data.frame
all_som_highest$som_highest <- all_som_highest$highest

all_fpn_highest <- lapply(unique(all_tc$MRI_S_ID), function(x) net_rank(sesh=x, network="Frontoparietal", all_data=all_tc)) %>% do.call(rbind,.) %>% as.data.frame
all_fpn_highest$fpn_highest <-all_fpn_highest$highest 


```


### Load sistat data and get lists of scans to use
All sistat tables should be exported as CSVs into a single directory
the next several chunks deal with reading, cleaning and annotating the data exported from sistat, and then age matching
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
# set location of directory with ucla sistat CSVs
csvdir_ucla <- file.path(project,"demographics/ucla_sistat")

# get list of files_ucla in directory
files_ucla <- list.files(csvdir_ucla)
fpaths <- lapply(files_ucla, function(file) paste(csvdir_ucla,file,sep="/"))

# clean names
fnames <- gsub(".csv","",files_ucla)
fnames <- gsub("Re22Q_","",fnames)
fnames <- gsub("Form_","",fnames)
fnames <- gsub("Qry_","",fnames)

# read all, set to na: "-9999", "-9998","." 
input_all_ucla <- lapply(fpaths, read.csv, header=T, na.strings=c(".","-9999","-9998"), strip.white=T, sep=",")
names(input_all_ucla) <- fnames
df_all_ucla <- lapply(input_all_ucla, function(x) data.frame(x))
ucla_demo <- df_all_ucla$demo_mri

# remove "FAMILY MEMBER" designation from subject identity
ucla_demo$SUBJECT_IDENTITY <- ucla_demo$SUBJECT_IDENTITY %>% sub("FAMILY MEMBER","",.) %>% sub(",","",.) %>% trimws(which="both") %>% as.factor
# change sex coding from 0/1 to F/M and set to factor
ucla_demo$SEX <- factor(ucla_demo$SEX,levels=c(0,1),labels=c("F","M"))
```

temporary: read temporary csv with several subjects not yet in sistat
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
# TODO: this chunk is temporary until sistat is updated 
# TODO: note: q_0526 sex was "na" in original sheet, changed to M because combat can't have NAs
# read new data
temp_demo <- read_xlsx(file.path(project,"demographics/temporary/sMRI_demo_info_forCharlie.xlsx"), col_names=TRUE,na="",trim_ws = TRUE)

# make empty demographics data frame to add new data to
demo_add <- ucla_demo[1:nrow(temp_demo),]
demo_add[,] <- NA
demo_add$SUBJECTID <- temp_demo$`Subject ID`
demo_add$SUBJECT_IDENTITY <- temp_demo$Diagnosis
demo_add$MRI_S_ID <- temp_demo$`MRI ID`
demo_add$SEX <- as.factor(temp_demo$Sex)
demo_add$AGE <- temp_demo$Age
demo_add$AGEMONTH <- temp_demo$Age*12
demo_add$CONVERTEDVISITNUM <- 2

# append to ucla demo
ucla_demo <- rbind(ucla_demo,demo_add)
```
 
continue regular steps
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
# subset demo_mri for used scans
ucla_demo <- filter(ucla_demo, MRI_S_ID %in% all_sessions)

# manually fix missing sex for Q_0381_09102019
# TODO: fix in sistat and re-export
ucla_demo[which(ucla_demo$MRI_S_ID == "Q_0381_09102019"),"SEX"] <- "F"

# set race=NA to 7 (unknown)
ucla_demo$RACE[is.na(ucla_demo$RACE)] <- 7
# set race as factor 1=American Indian/Alaska Native; 2=Asian; 3=Native Hawaiian/Pacific Islander; 4=Black or African American; 5=White; 6=Multiple; 7=Unknown
ucla_demo$RACE <- factor(ucla_demo$RACE,levels=c(1:7),labels=c("1_Native_American","2_Asian","3_Pacific_Island","4_Black","5_White","6_Multiple","7_Unknown"))
# ethnicity as factor with 0=N 1=Y
ucla_demo$HISPANIC[is.na(ucla_demo$HISPANIC)] <- "Unknown"
ucla_demo$HISPANIC <- factor(ucla_demo$HISPANIC,levels=c(0,1,"Unknown"),labels=c("N","Y","Unknown"))
# get more accurate age with AGEMONTH/12
ucla_demo$AGE <- as.numeric(ucla_demo$AGEMONTH)/12 

# subset to under max age limit (23 years old)
ucla_demo_agelim <- filter(ucla_demo, ucla_demo$AGE <= 23)

# subset to hcs del
ucla_demo_agelim_group <- filter(ucla_demo_agelim, SUBJECT_IDENTITY=="CONTROL" | SUBJECT_IDENTITY =="PATIENT-DEL")

# remove unused factor levels
ucla_demo_agelim_group %<>% droplevels

# function to add column to code timepoints relative to sample used (i.e. if visit 1 and 1.12 missing, then 1.24 is baseline)
# trio/prisma coded as T/P-visit_n where T-1 would be the subject's first trio scan and P-1 the first prisma, P-2 the second...
# function should be applied to the indicies of rows (r) in a subset of demo_mri
gettp <- function(r, df){
  sub <- df$SUBJECTID[[r]]
  visit <- df$CONVERTEDVISITNUM[[r]]
  all_visits <- df$CONVERTEDVISITNUM[which(df$SUBJECTID == sub)] %>% sort
  n_visits <- length(all_visits)
  nt_visits <-length(which(all_visits < 2))
  np_visits <- length(which(all_visits >= 2))
  visit_index <- which(all_visits == visit)
  if (visit < 2){
    label=paste("T-",visit_index,sep="")
  }else if (visit >= 2){
    p_visits <- all_visits[which(all_visits >= 2)] %>% sort
    p_visit_index <- which(p_visits == visit)
    label=paste("P-",p_visit_index,sep="")
  }
  return(data.frame(SUBJECTID=sub,CONVERTEDVISITNUM=visit,converted_timepoint=label,n_timepoints=n_visits,n_trio=nt_visits,n_prisma=np_visits,visit_index=visit_index))
}

# get timepoints
#timepoints <- sapply(1:nrow(ucla_demo_agelim_group),function(r) gettp(r,ucla_demo_agelim_group)) %>% t %>% as.data.frame
timepoints <- lapply(1:nrow(ucla_demo_agelim_group),function(r) gettp(r,ucla_demo_agelim_group)) %>% do.call(rbind,.) %>% as.data.frame

#colnames(timepoints) <- c("SUBJECTID","CONVERTEDVISITNUM","converted_timepoint","n_timepoints","n_trio","n_prisma","visit_index")
#ucla_demo_hcs_del <- cbind(ucla_demo_agelim_group,timepoints[,3:7])
ucla_demo_hcs_del <- merge(x=ucla_demo_agelim_group,y=timepoints, by=c("SUBJECTID","CONVERTEDVISITNUM"))

ucla_demo_hcs_del$visit_index %<>% as.factor

```

All timepoints, pre-matching demographics summary
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
demo_summary <- CreateTableOne(data=ucla_demo_hcs_del,vars=c("AGE","SEX"),strata="SUBJECT_IDENTITY",addOverall=F)
print(demo_summary, showAllLevels=T)
```

Baseline pre-matching summary
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
demo_summary_bl <- CreateTableOne(data=filter(ucla_demo_hcs_del, ucla_demo_hcs_del$visit_index == 1),vars=c("AGE","SEX"),strata="SUBJECT_IDENTITY",addOverall=F)
print(demo_summary_bl)
```



Merge TC with demo_mri
```{r message=FALSE, warning=FALSE, paged.print=FALSE}

demo_mri_tc <- merge(x=ucla_demo_hcs_del, y=all_fpn_highest[,c("MRI_S_ID","fpn_highest")], by="MRI_S_ID")
demo_mri_tc <- merge(x=demo_mri_tc, y=all_som_highest[,c("MRI_S_ID","som_highest")], by="MRI_S_ID")
demo_mri_tc_hcs_del <- demo_mri_tc %>% filter(SUBJECT_IDENTITY=="CONTROL" | SUBJECT_IDENTITY =="PATIENT-DEL")

# change class to factor
demo_mri_tc_hcs_del$SUBJECT_IDENTITY <- factor(demo_mri_tc_hcs_del$SUBJECT_IDENTITY)
demo_mri_tc_hcs_del$SUBJECTID <- factor(demo_mri_tc_hcs_del$SUBJECTID)
#demo_mri_tc_hcs_del$site <- factor(demo_mri_tc_hcs_del$site)
demo_mri_tc_hcs_del$MRI_S_ID <- factor(demo_mri_tc_hcs_del$MRI_S_ID)

# create column for age over 14
demo_mri_tc_hcs_del$over_13 <- demo_mri_tc_hcs_del$AGE > 13
demo_mri_tc_hcs_del$over_13 <- gsub(TRUE,"over_13",demo_mri_tc_hcs_del$over_13)

# rename highest columns
demo_mri_tc_hcs_del$fpn_highest <- gsub(TRUE, "fpn_highest", demo_mri_tc_hcs_del$fpn_highest)
demo_mri_tc_hcs_del$som_highest <- gsub(TRUE, "som_highest", demo_mri_tc_hcs_del$som_highest)

# subset to baseline
demo_mri_tc_hcs_del_bl <- filter(demo_mri_tc_hcs_del, visit_index==1)

# create contingency tables
cont_som <- table(demo_mri_tc_hcs_del_bl$over_13, demo_mri_tc_hcs_del_bl$som_highest)

# create contingency tables
cont_fpn <- table(demo_mri_tc_hcs_del_bl$over_13, demo_mri_tc_hcs_del_bl$fpn_highest)

# chi squared
chisq_som <- chisq.test(cont_som)
chisq_fpn <- chisq.test(cont_fpn)

```


